---
title: "Exercise 2A: Visium HD QC"
format: html
editor: source
editor_options: 
  chunk_output_type: console
execute:
  warning: false
---


## Quality control at cell-level

In this second exercise, we will focus on the critical step of quality control (QC) for sequencing-based spatial transcriptomics data. We will focus on the segmented Visium HD dataset used in `Exercise 1C`, so each observation is an individual cell with the selected region of interest. QC helps flag low-quality cells or technical artifacts before downstream analysis.

::: callout-note
A similar QC can be done per bin or aggregated bin if a binned analysis of Visium HD is performed.
:::

## Learning objectives

By the end of this exercise, you will be able to:

- Calculate per-cell QC metrics.
- Identify local outliers based on various QC metrics.
- Detect spatial artifacts using `SpotSweeper`.
- Visualize QC metrics and detected artifacts.


## Libraries
```{r day1-2a-spotsweeper-qc-segmented-1}
#| message: false
#| warning: false
#| output: false

library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(Voyager)
library(HDF5Array)
library(alabaster.sfe)

library(SpotSweeper)
library(scuttle)
library(scrapper)
library(scater)
library(escheR) 

library(ggplot2)
library(ggside)
library(ggspavis)
library(patchwork)
```

## Calculate QC metrics

We will start by calculating QC metrics that are also commonly used in scRNA-seq data analysis: total UMI counts, number of detected genes, and percentage of UMIs originating from genes of the mitochondrial genome. Here these metrics are calculated per segmented cell and used to identify low-quality cells.

We will start by loading our `SpatialFeatureExperiment` object, which was saved in the previous exercise, and prepare it for quality control analysis.  


```{r day1-2a-spotsweeper-qc-segmented-2}
# Reload the SpatialFeatureExperiment object if not in the R session already
sfe <- readObject("results/day1/01.1c_sfe_visium/")
## Otherwise, quickly go through Exercise day1-1c to recreate the object

# Identify mitochondrial genes (genes with symbol starting with "MT-")
is.mito <- rownames(sfe)[grepl("^MT-", rownames(sfe))]
```

<!--- NOTE: HDF5-backed saving/loading via saveHDF5SummarizedExperiment() is reliable for assays, but it can leave sf geometries with invalid external pointers after reload (GEOS/s2), which then crash functions like localOutliers() with “external pointer is not valid”. 
Need a solution or don't delete previous object 
--->

Now, we will calculate per-cell QC metrics using the `scuttle` package. The function `addPerCellQCMetrics` adds QC metrics, including mitochondrial gene percentage, to the `colData` of the `SpatialFeatureExperiment` object. 

::: callout-note
This function will soon be deprecated and replaced by a faster function in the `scrapper` package (function `quickRnaQc.se()`).
:::

Run the following code to compute these metrics:

```{r day1-2a-spotsweeper-qc-segmented-3}
sfe <- addPerCellQCMetrics(sfe, subsets = list(Mito = is.mito))
```

::: callout-important
## Exercise 1

Check which metadata has been added to `colData`. Do you recognize the different metrics?
:::

:::{.callout-tip collapse="true"}
### Answer

```{r day1-2a-spotsweeper-qc-segmented-4}
# Display the colData to see the newly added QC metrics.
colData(sfe)
```

- The `sum` column contains the total number of unique molecular identifiers (UMIs) per cell
- The `detected` column contains the number of unique genes detected per cell
- The `subsets_Mito_percent` column contains the percentage of transcripts mapping to mitochondrial genes per cell
:::

### Global outlier detection

In order to detect low-quality cells and filter them out, QC methods similar to those from scRNA-seq analysis can be applied. A simple approach is to apply a fixed upper or lower threshold to a metric across all cells and remove the cells that do not pass the filtering criteria.

To set up these thresholds and adapt them to this technology, we can look for information from 10x Genomics and the community, for example https://github.com/10XGenomics/HumanColonCancer_VisiumHD/issues/28. Based on this discussion the authors of the OSTA book wrote (see "Remove bins overlaying empty tissue" section in https://lmweber.org/OSTA/pages/seq-workflow-visium-hd-bin.html#dependencies):

> Based on a discussion with the original author at 10x Genomics, in this dataset, 8 µm bins with a library size below 100 are removed. Since we expect a 4-fold increase in library size across all bins at 16 µm, we can set the filtering threshold to 400.

This is not exactly transposable to the segmented cell output, but conveys the message that the UMI counts are low, much lower than what is observed with scRNA-seq data! We can also have a look and plot distributions of the values to decide which is the best cutoff:

```{r day1-2a-spotsweeper-qc-segmented-5}
# Plot library size vs. detected genes
p1 <- ggcells(sfe, aes(x=detected, y=sum)) + 
  geom_point() +
  geom_density_2d() +
  scale_x_log10() +
  scale_y_log10() +
  geom_xsidehistogram() + 
  geom_ysidehistogram() + 
  geom_vline(xintercept = 200, col="red", lty=2) +
  geom_hline(yintercept = 300, col="red", lty=2) +
  labs(y="Number of UMIs", x="Number of detected genes") +
  ggtitle("Library size vs. detected")

# Plot mito proportion vs. detected genes
p2 <- ggcells(sfe, aes(x=detected, y=subsets_Mito_percent)) + 
  geom_point() +
  geom_density_2d() +
  scale_x_log10() +
  geom_xsidehistogram() + 
  geom_ysidehistogram() + 
  geom_vline(xintercept = 200, col="red", lty=2) +
  geom_hline(yintercept = 10, col="red", lty=2) +
  labs(y="% mitochondrial reads", x="Number of detected genes") +
  ggtitle("Mito proportion vs. genes detected")

p1 + p2
```

And further check how many cells would be removed with the given thresholds:

```{r day1-2a-spotsweeper-qc-segmented-6}
# Flag cells based on fixed thresholds
sfe$qc_lib_size <- sfe$sum < 300
sfe$qc_detected <- sfe$detected < 200
sfe$qc_mito_prop <- sfe$subsets_Mito_percent > 10

# Tabulate number of cells flagged by each
qc <- grep("^qc", names(colData(sfe)))
sapply(colData(sfe)[qc], table)
```

::: callout-important
## Exercise 2
What do you think about this potential QC filtering?

Have a look at the metrics on the tissue section using the plotting functions used in the previous practical, or with the `plotObsQC()` function from the `ggspavis` package. Where are the cells that would be excluded with such global cutoffs? 

Try using more stringent cutoffs to see which cells would be excluded. What do you notice? What does it tell us retrospectively about the use of global cutoffs in the analysis of scRNA-seq data? 
:::

:::{.callout-tip collapse="true"}
### Answer

```{r day1-2a-spotsweeper-qc-segmented-7}
## Plot the metrics
p1 <- plotSpatialFeature(sfe, colGeometryName = "cellSeg", features = "subsets_Mito_percent") +
  theme(plot.title=element_text(hjust=0.5)) +
  scale_fill_gradientn(colors=pals::jet())
p2 <- plotSpatialFeature(sfe, colGeometryName = "cellSeg", features = "sum") +
  theme(plot.title=element_text(hjust=0.5)) +
  scale_fill_gradientn(trans = "log", colors=pals::jet())
p1 + p2
  
## Which cells would be excluded
p1 <- plotSpatialFeature(sfe, colGeometryName = "cellSeg", features = "qc_lib_size") +
  theme(plot.title=element_text(hjust=0.5)) +
  scale_fill_manual(values = c("beige", "red"))
p2 <- plotSpatialFeature(sfe, colGeometryName = "cellSeg", features = "qc_mito_prop") +
  theme(plot.title=element_text(hjust=0.5)) +
  scale_fill_manual(values = c("beige", "red"))
p1 + p2


## A similar plot using the ggspavis::plotVisium() function (y-coordinates are flipped!)
p1 <- plotVisium(sfe, annotate = "sum", image = FALSE) +
  theme(plot.title=element_text(hjust=0.5)) +
  scale_fill_gradientn(colors=pals::jet())
p2 <- plotVisium(sfe, annotate = "sum", image = FALSE) +
  theme(plot.title=element_text(hjust=0.5)) +
  scale_fill_gradientn(trans = "log", colors=pals::jet())
p1 + p2

## or with the ggspavis::plotObsQC() function
p1 <- plotObsQC(sfe, 
                plot_type="spot", annotate="qc_lib_size") + 
  ggtitle("Library size (< 300 UMI)")
p2 <- plotObsQC(sfe, 
                plot_type="spot", annotate="qc_mito_prop") + 
  ggtitle("Mito proportion (> 10%)")
p1 + p2

```

Stringent cutoffs lead to excluding cells in specific cellular structures, so quite likely excluding some specific cell types, which is not ideal. We have the opportunity to see that directly on the slide here, which is not the case when we analyze scRNA-seq data... There, a choice of a stringent cutoff can filter out entirely some cell types without us noticing!  


:::

### Local Outlier Detection

To avoid discarding whole layers of tissue, another approach to identify low-quality cells is to use local outlier detection methods that take into account the spatial context of each cell.

The `SpotSweeper` package provides functions to identify local outliers based on common QC metrics (such as the ones we have seen in the previous section: library size, number of detected genes, and mitochondrial percentage).

We will use `localOutliers` from `SpotSweeper`, which helps identify cells that deviate significantly from their local neighborhood.


```{r day1-2a-spotsweeper-qc-segmented-8}
# Identify local outliers based on library size ("sum" of counts).
# cells with unusually low library size compared to their neighbors will be flagged.
sfe <- localOutliers(sfe,
  metric = "sum",
  direction = "lower",
  log = TRUE
)

# Identify local outliers based on the number of unique genes detected ("detected").
# cells with an unusually low number of detected genes will be flagged.
sfe <- localOutliers(sfe, 
  metric = "detected",
  direction = "lower",
  log = TRUE
)

# Identify local outliers based on the mitochondrial gene percentage.
# cells with an unusually high mitochondrial percentage will be flagged.
sfe <- localOutliers(sfe,
  metric = "subsets_Mito_percent",
  direction = "higher",
  log = FALSE
)

# Combine all individual outlier flags into a single "local_outliers" column.
# A cell is considered a local outlier if it is flagged by any of the above metrics.
sfe$local_outliers <- as.logical(sfe$sum_outliers) |
  as.logical(sfe$detected_outliers) |
  as.logical(sfe$subsets_Mito_percent_outliers)
```

::: callout-important
## Exercise 3
How many cells were identified as local outliers based on the combined criteria?

As we have seen before, visualizing the QC metrics is crucial for understanding the quality of your spatial transcriptomics data and for deciding on appropriate filtering strategies, so look at their spatial localization using for example the `plotObsQC` function. What do you think?
:::


:::{.callout-tip collapse="true"}
### Answer
```{r day1-2a-spotsweeper-qc-segmented-9}
# Count the number of cells flagged as local outliers.
table(sfe$local_outliers)

plotObsQC(sfe, 
          plot_type="spot", annotate="local_outliers") + 
  ggtitle("All local outliers")
```

We can visualize the spatial distribution of the local outliers together with the QC metrics, for example plot the spatial distribution of local outliers for mitochondrial percentage.
```{r day1-2a-spotsweeper-qc-segmented-10}
plotQCmetrics(sfe,
  metric = "subsets_Mito_percent", outliers = "subsets_Mito_percent_outliers",
  point_size = 1, stroke = 0.5
) +
  ggtitle("Local outliers for %MT")
```

Overall we observe that very few outliers are detected. In regions where cells were spotted wiht global cutoffs, we notivce here that the cells that do not differ from their neighbourhood are not flagged anymore.
:::

::: callout-important
## Exercise 4
Plot the spatial distribution of local outliers for library size and number of detected genes. What do you observe?
:::

:::{.callout-tip collapse="true"}
### Answer

```{r day1-2a-spotsweeper-qc-segmented-11}
# Plot the spatial distribution of local outliers for library size.
plotQCmetrics(sfe,
  metric = "sum_log", outliers = "sum_outliers", 
  point_size = 1, stroke = 0.5
) +
  ggtitle("Local outliers for number of UMIs")


# Plot the spatial distribution of local outliers for number of detected genes.
plotQCmetrics(sfe,
  metric = "detected_log", outliers = "detected_outliers", point_size = 0.4,
  stroke = 0.75
) +
  ggtitle("Local outliers for number of detected genes")
```

Very few cells seem to be flagged as outliers
:::


We can visualize the z-scores from the local outlier detection test and compare them to the global threshold used earlier. How do you interpret this?

```{r day1-2a-spotsweeper-qc-segmented-12}
plotObsQC(sfe, plot_type = "violin", x_metric = "subsets_Mito_percent_z", annotate = "qc_mito_prop")
```

:::{.callout-tip collapse="true"}
### Answer
We notice that some cells that were labeled as outliers based on a global cutoff are not outliers anymore relative to their local neighborhood, and conversely.

:::

## Compare and combine global and local outliers and apply filtering

For the next steps we choose to:
- adjust the global cutoffs to be less stringent
- combine the (new) global and local outliers to perform the QC filtering

```{r day1-2a-spotsweeper-qc-segmented-13}
# adjust global filtering cutoff
sfe$qc_lib_size <- sfe$sum < 120
sfe$qc_detected <- sfe$detected < 100
sfe$qc_mito_prop <- sfe$subsets_Mito_percent > 30

# compare global and local outliers being flagged
sfe$global_outliers <- sfe$qc_lib_size | sfe$qc_detected | sfe$qc_mito_prop
p1 <- plotObsQC(sfe, plot_type="spot", annotate="global_outliers")
p2 <- plotObsQC(sfe, plot_type="spot", annotate="local_outliers")
gridExtra::grid.arrange(p1, p2, ncol=2)

# combine local and global outliers to create a final discard flag
sfe$discard <- sfe$global_outliers | sfe$local_outliers
```

::: callout-important
## Exercise 5
Visualize the cells that will be discarded based on the combined criteria. How many cells will be removed?
:::

:::{.callout-tip collapse="true"}
### Answer
```{r day1-2a-spotsweeper-qc-segmented-14}
# check how many cells will be discarded
table(sfe$discard)

# have a final look at the cells that we will discard
plotSpatialFeature(sfe, colGeometryName = "cellSeg", features = "discard")
```
:::

## Bonus: region-level technical artifacts

<!--- JUNA: maybe we could skip the technical artifacts part, the result is not easy to interpret in this section -->

Technical artifacts can arise during tissue preparation, for example tissue dissection can cause mechanical damage, leading to large areas with artificially low biological heterogeneity. These can be detected with the `findArtifacts` function of the `spotSweeper` package.

```{r day1-2a-spotsweeper-qc-segmented-15}
# find artifacts using SpotSweeper
sfe <- tryCatch(
  findArtifacts(sfe,
    mito_percent = "subsets_Mito_percent",
    mito_sum = "subsets_Mito_sum",
    n_order = 5,
    name = "artifact"
  ),
  error = function(e) {
    message(
      "findArtifacts() could not fit an artifact model for this subset: ",
      conditionMessage(e)
    )
    sfe$artifact <- FALSE
    sfe
  }
)
```

::: callout-important
## Exercise 6
What is the `findArtifacts` function performing and what are the (potentially problematic) assumptions?

Plot the spatial distribution of the `artifact` column identified by `SpotSweeper`. What do you observe?
:::

:::{.callout-tip collapse="true"}
### Answer

```{r day1-2a-spotsweeper-qc-segmented-16}
# Plot the spatial distribution of the detected artifacts.
plotQCmetrics(sfe,
    metric = "subsets_Mito_percent", 
    outliers = "artifact", 
    point_size = 1.1
) +
    ggtitle("Detected Spatial Artifacts")
```

The function is detecting region-level artifacts, such as "hangnail" artifacts, based on the multiscale variance of mitochondrial ratio. The exact methodology can be found in the paper and is actually quite complex and ad hoc (i.e., they devised this method after studying carefully the properties of their brain tissue section): https://pmc.ncbi.nlm.nih.gov/articles/PMC12258134/. 

Here indeed the method separates two global-regions, one on the left with less variability in %MT, and one on the right with more variability. Does this mean that the region on the left should be excluded here? Probably not...

:::

## Filtering 
We decide to filter out the low-quality cells from the `SpatialFeatureExperiment` object based on the combined criteria of global thresholds and local outliers. Additionally, we will remove any features (genes) that have zero counts across all remaining cells.

```{r day1-2a-spotsweeper-qc-segmented-17}
# remove the cells considered of low quality based on both global and local metrics
sfe <- sfe[, !sfe$discard]

# remove features no count at all
sfe <- sfe[rowSums(counts(sfe)) > 0, ]
sfe
``` 

What is the size of the dataset after filtering?

## Save the object

Save the filtered `SpatialFeatureExperiment` object for the next steps:
```{r day1-2a-spotsweeper-qc-segmented-18}
alabaster.sfe::saveObject(sfe, "results/day1/01.2_sfe_visium")
```

Clear your environment:
```{r day1-2a-spotsweeper-qc-segmented-19}
#| eval: false
rm(list = ls())
gc()
.rs.restartR()
```

:::{.callout-important}
**Key Takeaways:**

-   `scuttle::addPerCellQCMetrics` provides essential per-cell QC metrics.
-   `SpotSweeper::localOutliers` helps identify individual problematic cells relative to their local context.
-   `SpotSweeper::findArtifacts` is more controversial but could in certain case be useful to detect region-level artifacts
-   Visualizing QC results is critical for data interpretation and downstream analysis decisions.
:::
