---
title: "Data Preparation (2026 Course)"
format: html
editor: source
execute:
  eval: false
---

## Overview

This document describes the steps to download and prepare the datasets for the 2026 version of the course. We have shifted our focus to **Patient 2 (P2CRC)** from the Oliveira et al. (2025) study, utilizing both **Visium HD (Segmented)** and **Xenium** data.

## Official 10X Genomics Datasets

We will use the following datasets:

1.  **Visium HD (v4.0):** [Human Colon Cancer, CytAssist Gene Expression Libraries](https://www.10xgenomics.com/datasets/visium-hd-cytassist-gene-expression-libraries-of-human-crc-v4)
2.  **Xenium:** [Human Colon Cancer (P2CRC)](https://www.10xgenomics.com/platforms/xenium/product-family/dataset-human-crc)

## 1. Download Datasets

```{r download-data}
# Create data directory
if (!dir.exists("data")) dir.create("data")

# --- Visium HD (P2CRC) ---
# We download the 'Feature / Cell Matrix (Raw)' and 'Spatial Imaging Data' 
# or the full output folder if available. 
# Note: For the segmented output, we specifically need Space Ranger v4.0 outputs.

vhd_url <- "https://cf.10xgenomics.com/samples/spatial-exp/4.0.0/Visium_HD_Human_Colon_Cancer/Visium_HD_Human_Colon_Cancer_feature_slice.tar.gz"

if (!file.exists("data/Visium_HD_P2CRC.tar.gz")) {
  download.file(vhd_url, destfile = "data/Visium_HD_P2CRC.tar.gz")
  untar("data/Visium_HD_P2CRC.tar.gz", exdir = "data/P2CRC_VisiumHD")
}

# --- Xenium (P2CRC) ---
# We download the bundle containing the cell-feature matrix and spatial coordinates.

xen_url <- "https://cf.10xgenomics.com/samples/xenium/1.0.2/Xenium_V1_human_Colon_Cancer_P2_CRC/Xenium_V1_human_Colon_Cancer_P2_CRC_outs.tar.gz"

if (!file.exists("data/Xenium_P2CRC.tar.gz")) {
  download.file(xen_url, destfile = "data/Xenium_P2CRC.tar.gz")
  untar("data/Xenium_P2CRC.tar.gz", exdir = "data/P2CRC_Xenium")
}
```

## 2. Load and Inspect Objects

```{r load-data}
library(SpatialExperiment)
library(VisiumIO)
# library(XeniumIO) # Ensure this is installed

# Load Visium HD (Segmented)
spe_hd <- TENxVisiumHD(
  spacerangerOut = "data/P2CRC_VisiumHD/outs/",
  processing = "filtered",
  bin_size = "002" # or segmented
) |> import()

# Load Xenium
# spe_xen <- readXenium("data/P2CRC_Xenium/outs/")
```

## 3. Spatial Subsetting (TO DO)

The PDF requires subsetting both datasets to the **~same region** for side-by-side comparison. 

```{r subsetting-task}
# TO DO: Identify overlapping coordinates
# Example:
# spe_hd_sub <- spe_hd[, spatialCoords(spe_hd)[,1] > Xmin & ...]
# spe_xen_sub <- spe_xen[, spatialCoords(spe_xen)[,1] > Xmin & ...]

# Save the subsetted objects for Exercise 1
# saveRDS(spe_hd_sub, "data/spe_hd_P2CRC_sub.rds")
# saveRDS(spe_xen_sub, "data/spe_xen_P2CRC_sub.rds")
```
