---
title: "Exercise 1c: SpatialFeatureExperiment (SFE)"
format: html
editor: source
editor_options: 
  chunk_output_type: console
execute:
  warning: false
---

## Learning Objectives

By the end of this exercise, you will be able to:

-   Understand the structure of a `SpatialFeatureExperiment` (SFE) object.
-   Understand the differences between `SpatialExperiment` and `SpatialFeatureExperiment`.
-   Load Xenium data into an SFE object.
-   Access and visualize cell geometries and transcript spots.

## Libraries

```{r day1-1c-SFE-1}
#| message: false
#| warning: false

library(SpatialFeatureExperiment)
library(Voyager)
library(SFEData)
library(ggplot2)
library(patchwork)
library(SpatialExperiment)
```

## Introduction to SFE

`SpatialFeatureExperiment` (SFE) is an extension of `SpatialExperiment` designed to handle spatial data that includes more than just spot/cell coordinates. It is particularly powerful for imaging-based technologies like Xenium, as it can store:

-   **Cell boundaries** (as polygons/geometries).
-   **Transcript locations** (as point geometries).
-   **Tissue boundaries** or other annotations.

## 1. Load Xenium Data

We will use the Xenium data for **Patient 2 (P2CRC)** that was downloaded in the [Data Preparation](day1-0_data_preparation.qmd) step.

```{r day1-1c-SFE-2}
# Load pre-prepared P2CRC Xenium object (as SFE)
if (file.exists("../data/sfe_xen_P2CRC_sub.rds")) {
  sfe <- readRDS("../data/sfe_xen_P2CRC_sub.rds")
} else {
  message("P2CRC Xenium data not found! Please run day1-0_data_preparation.qmd first.")
  # Placeholder for loading from raw if needed:
  # sfe <- readXenium("data/P2CRC_Xenium/outs/", type = "sparse")
}
```

## 2. Exploring Geometries

One of the key features of SFE is the `colGeometries` slot, which stores geometries related to cells (columns).

::: callout-important
## Exercise 1

Check the available geometries in your SFE object using `colGeometries(sfe)`. What geometries are available?
:::

::: {.callout-tip collapse="true"}
## Answer
For Xenium data, you typically see `centroids` and `boundaries` (the cell polygons).

```{r day1-1c-SFE-3}
colGeometries(sfe)
```
:::

## 3. Visualization

SFE integrates with the `Voyager` package for high-level spatial plotting.

```{r day1-1c-SFE-4}
# Plot cell boundaries colored by a gene (e.g., EPCAM)
plotSpatialFeature(sfe, "EPCAM", colGeometryName = "boundaries")
```

::: callout-important
## Exercise 2

Compare the `SpatialExperiment` (SPE) and `SpatialFeatureExperiment` (SFE) representations of the same tissue region. What are the visual differences?
:::

::: {.callout-tip collapse="true"}
## Answer
The SPE object typically represents cells as points (spots), while the SFE object can represent the actual shapes of the cells obtained from segmentation. This is crucial for understanding cell morphology and precise transcript localization.
:::

## 4. Subsetting with Geometries

Since SFE uses the `sf` (simple features) framework, you can perform spatial subsetting using geometries.

```{r day1-1c-SFE-5}
# Example of subsetting to a specific area of interest (AOI)
# sfe_sub <- sfe[, st_intersects(colGeometry(sfe, "centroids"), aoi_geometry, sparse = FALSE)]
```

## Save the object

```{r day1-1c-SFE-6}
dir.create("results/day1", showWarnings = FALSE, recursive = TRUE)
saveRDS(sfe, "results/day1/01.1c_sfe_p2.rds")
```

::: callout-important
**Key Takeaways:**

-   `SpatialFeatureExperiment` is essential for imaging-based SRT with cell segmentation.
-   It stores complex geometries like cell boundaries and transcript spots.
-   `Voyager` provides powerful tools for visualizing and analyzing SFE objects.
:::
