---
title: "Exercise 1B: SpatialFeatureExperiment"
format: html
editor: source
editor_options:
  chunk_output_type: console
execute:
  warning: false
---

## Learning Objectives

By the end of this exercise, you will be able to:

-   Understand the structure of a `SpatialFeatureExperiment` object.
-   Explain when `SpatialFeatureExperiment` is useful compared with `SpatialExperiment`.
-   Access cell-level metadata and spatial coordinates from Xenium data.
-   Inspect geometry information such as cell or nucleus boundaries.
-   Compare the Xenium representation with the Visium HD representation from Exercise 1 A.

## Libraries

```{r day1-1b-SFE-1}
#| message: false
#| warning: false

library(SpatialFeatureExperiment)
library(SpatialExperiment)
library(Voyager)
library(ggplot2)
library(patchwork)
```

## Data for the course

We will work with the **Xenium In Situ, sample P2 CRC** dataset from the same human colorectal cancer study used in Exercise 1 A. The Xenium section is a serial section of the Visium HD P2 CRC sample.

This means the two datasets are biologically related but not the exact same physical tissue section. We therefore subset Xenium to an **approximately matching tissue region**, guided by the Visium HD region and by morphology. This makes the two exercises comparable while avoiding a full image-registration workflow during the practical.

The full Xenium output is large and includes cell boundaries, nucleus boundaries, transcript locations, morphology images, and expression matrices. For this exercise, we use a prepared course object generated by `data/prepare_2026_course_data.qmd`.

```{r day1-1b-SFE-2}
sfe_path <- "data/processed/sfe_xen_p2_crc_sub.rds"

if (!file.exists(sfe_path)) {
  stop(
    "The prepared Xenium object was not found. ",
    "Run data/prepare_2026_course_data.qmd first."
  )
}

sfe <- readRDS(sfe_path)
sfe
```

::: callout-important
## Exercise 1

Why is a `SpatialFeatureExperiment` useful for Xenium data?

Which information does Xenium provide that is not naturally represented by a simple spot- or bin-level `SpatialExperiment`?
:::

::: {.callout-tip collapse="true"}
## Answer

Xenium is an image-based spatial transcriptomics technology. In addition to a count matrix and cell coordinates, it can provide cell boundaries, nucleus boundaries, transcript locations, and morphology images.

`SpatialFeatureExperiment` extends `SpatialExperiment` with geometry support, making it better suited to storing and manipulating these spatial shapes.
:::

## Exploring the object

`SpatialFeatureExperiment` extends `SpatialExperiment`, so familiar accessors still apply:

```{r day1-1b-SFE-3}
dim(sfe)
colData(sfe) |> head()
rowData(sfe) |> head()
spatialCoords(sfe) |> head()
```

::: callout-important
## Exercise 2

Check the dimensions and metadata of the `sfe` object.

Questions:

-   What do the columns represent?
-   What do the rows represent?
-   How is this different from the Visium HD `spe` object in Exercise 1 A?
:::

::: {.callout-tip collapse="true"}
## Answer

In this object, columns represent Xenium cells, while rows represent measured genes/features. This differs from the Visium HD binned `SpatialExperiment`, where columns represent spatial bins rather than segmented cells.

```{r day1-1b-SFE-4}
dim(sfe)
ncol(sfe)
nrow(sfe)
```
:::

## Inspecting geometries

One of the key features of `SpatialFeatureExperiment` is that it can store geometries associated with cells or features.

```{r day1-1b-SFE-5}
colGeometries(sfe)
```

::: callout-important
## Exercise 3

Inspect the available column geometries in the Xenium object.

Which geometries are available? What biological structures do they correspond to?
:::

::: {.callout-tip collapse="true"}
## Answer

For Xenium data, the available geometries commonly include cell centroids, cell boundaries, and nucleus boundaries. The exact names depend on the reader and package versions.

```{r day1-1b-SFE-6}
colGeometries(sfe)
```
:::

## Visualizing Xenium cells

We can start with a simple coordinate plot of the selected Xenium region:

```{r day1-1b-SFE-7}
xy <- as.data.frame(spatialCoords(sfe))
names(xy)[1:2] <- c("x", "y")

ggplot(xy, aes(x = x, y = y)) +
  geom_point(size = 0.08, alpha = 0.4) +
  coord_fixed() +
  theme_bw() +
  labs(x = "x", y = "y")
```

::: callout-important
## Exercise 4

Compare this point-based visualization with the Visium HD plot from Exercise 1 A.

What does one point represent in each dataset?
:::

::: {.callout-tip collapse="true"}
## Answer

In the Visium HD exercise, each point or square represents a 16 um spatial bin. In the Xenium exercise, each point represents a segmented cell centroid. This is one of the key conceptual differences between binned sequencing-based spatial data and image-based cell-resolved spatial data.
:::

We can also visualize expression for a marker gene. For example, `EPCAM` is often informative in epithelial tumor regions:

```{r day1-1b-SFE-8}
plotSpatialFeature(sfe, "EPCAM")
```

::: callout-important
## Exercise 5

Choose another marker gene and visualize its spatial pattern.

How does the interpretation differ from plotting a gene on Visium HD bins?
:::

::: {.callout-tip collapse="true"}
## Answer

The Xenium signal is measured over segmented cells, while the Visium HD binned signal aggregates molecules within fixed spatial bins. Xenium can therefore represent cell-level heterogeneity more directly, but it measures a targeted panel rather than the whole transcriptome.
:::

## Subsetting with coordinates

The prepared object has already been subsetted during data preparation. We can still apply coordinate-based subsetting in the same spirit:

```{r day1-1b-SFE-9}
xy <- spatialCoords(sfe)

roi <- c(
  xmin = quantile(xy[, 1], 0.25),
  xmax = quantile(xy[, 1], 0.75),
  ymin = quantile(xy[, 2], 0.25),
  ymax = quantile(xy[, 2], 0.75)
)

keep <- xy[, 1] > roi[["xmin"]] & xy[, 1] < roi[["xmax"]] &
  xy[, 2] > roi[["ymin"]] & xy[, 2] < roi[["ymax"]]

sfe_sub <- sfe[, keep]
dim(sfe)
dim(sfe_sub)
```

::: callout-important
## Exercise 6

Why did we subset Xenium during data preparation instead of using the full public output directly in the exercise?
:::

::: {.callout-tip collapse="true"}
## Answer

The full Xenium output is large and contains many cells, boundaries, transcript coordinates, and morphology images. A course-sized subset keeps the exercise responsive while preserving the structure of a real Xenium object.

The subset is also chosen to approximately match the Visium HD P2 CRC region from Exercise 1 A. Because Visium HD and Xenium are serial sections, this is a morphology-matched comparison rather than an exact same-cell comparison.
:::

## Save the object

```{r day1-1b-SFE-10}
dir.create("results/day1", showWarnings = FALSE, recursive = TRUE)
saveRDS(sfe, "results/day1/01.1b_sfe_p2.rds")
```

Clear your environment:

```{r day1-1b-SFE-11}
#| eval: false
rm(list = ls())
gc()
.rs.restartR()
```

::: callout-important
**Key Takeaways:**

-   `SpatialFeatureExperiment` is useful for geometry-rich spatial data.
-   Xenium observations are cell-resolved, while the Visium HD object in Exercise 1 A is bin-resolved.
-   The P2 CRC subsets are designed to be comparable in course scope and approximate tissue region, not identical at cell level.
:::
