---
title: "Exercise 8 : Spatial Stats"
editor: source
editor_options: 
  chunk_output_type: console
bibliography: ../references.bib
format:
  html:
    embed-resources: true
---

## Learning objectives

- Learn the basic principles of exploratory spatial statistics
- Compute and visualise local and global measures of spatial autocorrelation
- Compute and interpret point pattern analysis metrics on cell-based data
- Compute and interpret geometric measures of multicellular structures
- Compare spatial statistics measures across samples using linear mixed models
- This chapter is based on the `OSTA` online book chapter on spatial statistics and uses heavily the `Voyager` spatial statistics library [@moses2023voyager]

## Libraries
```{r day3-1-spatial-stats-1}
#| message: false
#| warning: false
#| output: false

library("SpatialExperiment")
library("ggplot2")
library("Voyager")
library("HDF5Array")
library("tidyr")
library("dplyr")
library("lme4")
library("lmerTest")
library("spatialFDA")
library("scrapper")
library("dittoSeq")
library("sosta")
library("alabaster.sfe")
set.seed(1234)
```

```{r day3-1-spatial-stats-2}
knitr::opts_chunk$set(cache.lazy = FALSE)
```

## Load data

```{r day3-1-spatial-stats-3}
# Load the SpatialExperiment objectq
sfe <- readObject("results/day2/02.4_sfe/")
# set imageData to NULL as we do not need it anymore
imgData(sfe) <- NULL

# Let's define the color palette for plotting
cls <- levels(factor(colData(sfe)$SingleR_label))
ctc <- setNames(unname(dittoColors())[seq_along(cls)], cls)
```

In spatial statistics, we often compare categorical labels across space. Useful categorical labels for spatial omics datasets include non-spatial clusters or cell types [@aran2019reference].

## Data modalities of spatial statistics

In spatial transcriptomics, we differentiate between imaging-based and sequencing-based technologies, which give rise to very different _data modalities_ [@rao2021exploring].  

![Data Modalities in Spatial Transcriptomics [@emons2025harnessing]](https://oup.silverchair-cdn.com/oup/backfile/Content_public/Journal/nar/53/17/10.1093_nar_gkaf870/1/gkaf870fig1.jpeg?Expires=1783069626&Signature=VfVM2TBfNTmXSzsf6VyjrQAfQeWfeqIIR6L4DXj-npV22XVqy5VlrzXm9lqR4aYGtdeK4L40q04YeOeimaWAFYdLBkN2jAO9BfRry7FEGUUYGzoYqvS7kyNLFhMGjhisWUfJO~ZZ24DT8FFLSsp3Sp6HR0Zi6q9s32Hx7ZzJGlRmV9vPN-H8yHUyDM9WUy-MTZY~406LScggMkd4DjhPLldWcwjDG10CEeDcRb3qjI-q3zBk63sZLOJFiFPcH8EGyj6UMi-STQf5dnvzPAqS6AZ6ryYDOS7cPIrG0tUbrIlZSDFLL8H298XM0Q-rI9fbqZe42AScZ9vNWD1xGZo99w__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA)

Our dataset was acquired using 10x Visium HD. The data modality of this technology is best described by a _regular lattice_, which means that the observations are taken at regularly spaced intervals. This is very different to imaging-based approaches where observations can be thought of as a stochastic data generating process that is related to some underlying biological mechanism. 

With spots of 2um^2, Visium HD can provide subcellular resolution. We can segment cells from this regular lattice of pixels and approximate the cells by their centroid. This in turn can be described as an _irregular lattice_ of the cell types in space or even be analysed as a _point patterns_ using the cell centroid locations. 

In what follows, we will perform three types of analyses on different _spatial scales_ of the dataset. 

- i) cell-free analysis: This is an irregular lattice and we analyse it thus with tools from lattice data analysis 
- ii) cell-level analysis: This is a point pattern pattern of cell centroids and we will analyse it with point pattern analysis.
- iii) multi-cellular analysis: This is a structure-based analysis that we will analyse with geometric tools.

## Cell type-free analysis

Lattice-based analysis is focused on comparing a numerical value across the lattice of measurements. Such a lattice is constructed by defining the neighbourhood of each cell. First, we will define a neighbourhood on which we want to compute spatial statistics metrics. In 10x Visium, it is very natural to consider the direct neigbours of hexagonal visium lattice. We do this with the function `Voyager::findVisiumGraph()`. 

```{r day3-1-spatial-stats-4}
#| fig-width: 7
#| fig-height: 7
(g <- findSpatialNeighbors(sfe,
    method = "knearneigh", # wraps the spdep function with the same name
    k = 5,
    zero.policy = TRUE,
    sample_id = "all"
  ))

colGraph(sfe, "visium") <- g

plotColGraph(sfe,
  colGraphName = "visium"
) + theme_void()
```

We notice two things:

- The $y$-coordinate is flipped compared to the image. This is due to the way that visium data is acquired. The Visium slide is placed on top of the tissue but the H&E image is acquired from below the tissue.

- On top of the spots, we now see the hexagonal neighbourhood on top of the regular lattice. 

### Global indicators of spatial association

Now that we have defined both the neighbourhood of a spot and preprocessed the gene expression, we can turn to measures of spatial association. 

In general, in a global association metric, we compute for each point, $i$, a metric for its neighbourhood $j$. The neighbourhood we quantified before as the regular hexagonal grid and is passed to the functions as a weight matrix $w_{ij}$. The association measure is then a function $f(x_i,x_j)$ and could e.g. quantify spatial correlation

$$
\sum_i \sum_j f(x_i,x_j) w_{ij}
$$

A famous measure of global spatial autocorrelation is called Moran's $I$, which assesses correlation of a gene with itself given a defined weight matrix $w$ [@moran1950notes]. 

::: callout-important
### Exercise

- Extract the top 200 highly variable genes in this dataset with `scrapper::modelGeneVariances()` and `scrapper::chooseHighlyVariableGenes`
- Calculate the global Moran's $I$ coefficient with the function `Voyager::runUnivariate()` on these 200 highly variable genes. You can find information and examples on how to run this function by typing `?runUnivariate` in your console
- Extract the top 3 genes with the highest Moran's $I$ value
- Plot the expression of these three genes in space using `Voyager::plotSpatialFeature`

:::

::: {.callout-tip collapse="true"}
### Answer

First, we model the variance on a subset of the genes for computational reasons
Then, we extract the 200 most highly variable genes

```{r day3-1-spatial-stats-5}
gene.var <- modelGeneVariances(logcounts(sfe), 
                               num.threads = 1)
# chooseHighlyVariableGenes() expects the residual variance statistics
hvg.index <- chooseHighlyVariableGenes(gene.var$statistics$residuals, 
                                      top = 100)

#add the hvgs to the `sfe` - adapted from https://github.com/libscran/scrapper/blob/master/R/se_chooseRnaHvgs.R
is.hvg <- logical(nrow(gene.var$statistics))
is.hvg[hvg.index] <- TRUE
gene.var[["hvg"]] <- is.hvg
rowData(sfe) <- cbind(rowData(sfe), gene.var)
#extract the gene names of the hvgs
hvg <- rownames(sfe)[is.hvg]
```

Next, we calculate global Moran's I for these 200 genes

```{r day3-1-spatial-stats-6}
sfe <- runUnivariate(sfe, 
    type="moran", 
    features=hvg, 
    colGraphName="visium")
```

We order them and plot the gene expression top 3 genes wiht the `plotSpatialFeature` function from the `Voyager` package.

```{r day3-1-spatial-stats-7}
I <- rowData(sfe)$moran_sample01
o <- order(I, decreasing=TRUE)
topGenes <- rownames(sfe)[head(o, 3)]
plotSpatialFeature(sfe, topGenes, ncol=3)
```

::: 

### Local indicators of spatial association

In the section above, we have computed a global measure of spatial association, meaning that we get one number per field-of-view. This is an average of the local contributions and neglects the potential underlying heterogeneity completely. 

Therefore, we will investigate _local_ indicators of spatial association [@anselin1995local]. 

In the case of Moran's $I$, we will now not calculate one measure for the entire field-of-view but rather one metric per location/spot $i$: $I_i$.

::: callout-important
## Exercise

- Calculate local Moran's $I$ for the top three global Moran's $I$ genes from above using the function `Voyager::runUnivariate()`

- Plot the local Moran's $I$ values in space using `Voyager::plotLocalResult`

:::

::: {.callout-tip collapse="true"}
### Answer

First, we have to calculate the metric on the first most autocorrelated genes. 
```{r day3-1-spatial-stats-8}
sfe <- runUnivariate(sfe,
                     features = topGenes,
                     colGraphName = "visium",
                     type = "localmoran")
```

Then, we plot the local Moran's $I$ values in space. 

```{r day3-1-spatial-stats-9}
#| fig-width: 7
#| fig-height: 15
plotLocalResult(sfe, "localmoran",
                features =topGenes, 
                ncol = 1,
                colGeometryName = "centroids",
                divergent = TRUE, 
                diverge_center = 0)
```

:::

### Discretising regions with Moran's scatterplot

The continuous local Moran's $I_i$ measurements can be difficult to interpret. A simplification is Moran's scatter plot [@anselin2019moran]. The idea is to compare the Moran's $I$ of the spot $i$ itself with its neighbours $j$. Like this we obtain four options: high $I_i$ next to neighbours that have high $I_j$, high $I_i$ next to neighbours that have low $I_j$ etc. This separates outliers (high-low and low-high) from values within more homogeneous regions (high-high and low-low).

::: callout-important
### Bonus Exercise

- Calculate Moran's scatterplot with `runUnivariate` on the top three global Moran's $I$ genes from above

- Plot the Moran's scatterplot results in space using `Voyager::plotLocalResult`

:::

::: {.callout-tip collapse="true"}
### Answer

First, we have to calculate the metric on the first three most variable HVGs. 
```{r day3-1-spatial-stats-10}
sfe <- runUnivariate(
    sfe,
    features =  topGenes,
    colGraphName = "visium",
    type = "moran.plot"
  )
```

We will show the Moran's scatter plot for the gene `PIGR`.

```{r day3-1-spatial-stats-11}
# we will show one Moran's scatterplot
moranPlot(sfe, topGenes[1], graphName = "visium", swap_rownames = "symbol") + theme_classic()
```

The grey dashed lines indicate the mean expression of PIGR in the spots themselves and among their neighbours. According to these means, the spots are categorised into 4 categories, which can be plotted back in space. We notice that there is an interesting part of the tissue that constitutes as "high-high" for all three genes in the same region (bottom left).

```{r day3-1-spatial-stats-12}
#| fig-width: 7
#| fig-height: 15
plotLocalResult(
    sfe,
    name = "localmoran",
    features = topGenes,
    ncol = 1,
    attribute = "mean",
    colGeometryName = "centroids"
)
```

We can also filter on the p-values for non-significant cells 

```{r day3-1-spatial-stats-13}
localResults(sfe)[["localmoran"]][["PIGR"]]<- 
  localResults(sfe)[["localmoran"]][["PIGR"]] |>
  mutate(locClust = ifelse(`-log10p_adj` > -log10(0.05), as.character(mean), "non-siginificant"))
```

```{r day3-1-spatial-stats-14}
#| fig-width: 7
#| fig-height: 7
plotLocalResult(
    sfe,
    name = "localmoran",
    features = "PIGR",
    ncol = 1,
    attribute = "locClust",
    colGeometryName = "centroids"
)
```

:::

## Cell type-based analysis

In the cell-based analysis chapter, we will compare the distribution of categorical cell types in space using point pattern analysis. Let us first plot the cell types in space

```{r day3-1-spatial-stats-15}
#| fig-width: 10
#| fig-height: 10
plotSpatialFeature(sfe, 
    features="SingleR_label",
    colGeometryName="centroids",
    size = 1) +
    guides(col=guide_legend(override.aes=list(size=2))) +
    scale_color_manual(values = ctc)
```

### Point pattern analysis

Next, we will compute a measure of _spatial correlation across scales_ for all pairwise cell type combinations called Ripley's cross $K$ function using the R/Bioconductor package `spatialFDA` [@emons2026differential; @emons2025harnessing]. Ripley's cross $K$ function is defined as [@baddeley2016spatial]. 

$$
K(r) = \frac{1}{\lambda_j} \mathbb{E} [t(u,r,X^{j}) \mid u \in X^{i}],
$$

Ripley's cross $K$ function can be variance stabilised, which is then called Besag's cross $L$ [@baddeley2016spatial]. This is the measure we will compute for our dataset. 

$$
L(r) = \sqrt{\frac{K(r)}{\pi}}
$$

We will focus on three cell types to make the results more interpretable

```{r day3-1-spatial-stats-16}
resCross <- calcCrossMetricPerFov(
  sfe,
  correction = "iso",
  selection = c("Tumor", "T cells", "B cells"),
  subsetby = "sample_id",
  fun = "Lcross", 
  marks = "SingleR_label",
  rSeq = seq(0, 100, length.out = 100), by = c("sample_id"),
  ncores = 1
)
```

```{r day3-1-spatial-stats-17}
#| fig-width: 10
#| fig-height: 7
plotCrossMetricPerFov(
  resCross,
  theo = TRUE,
  correction = "iso",
  x = "r",
  imageId = 'sample_id',
  linewidth = 1
) 
```

::: callout-important
### Exercise
What do you learn from these plots in combination with the distribution of cell types in space above? 
:::

::: {.callout-tip collapse="true"}
### Answer
We note a few things:

- All cells cluster among themselves and across scales (the diagonal of the plot)
- All cells show spacing (below the dashed line) of radii $r$ smaller than 15 $\mu$m $\rightarrow$ cell diameter
- Immune cells build aggregates
- Tumor cells show spacing in comparison to immune cells, there seems to be no immune invasion into the tumor
:::

## Multicellular analysis

### Reconstruction of the Tumor region

Next, we will continue with multicellular analysis. The R/Bioconductor package `sosta` is designed for structure-based analysis of multicellular anatomical structures. For example in spatial omics, "structures" could be germinal centers (GCs) in the tonsil, metastases or tertiary lymphoid structures (TLS) in cancer [@gunz2025analysis].

In `sosta`, the reconstruction of anatomical structures is based on the density profile of a marked point pattern. Based on an estimated threshold of the density profile the structures are segmented.

First, we will have a look at the point pattern intensity of the mark of interest. The histogram shows the distribution of intensity values of the pixel image on the left.

```{r day3-1-spatial-stats-18}
# Some of the later functions need to have an colData column with a image name identifier.
# We will define this here.
sfe$imageName <- "image1"

# Let us first look
shapeIntensityImage(
    sfe,
    marks = "SingleR_label",
    markSelect = "Tumor"
)
```

We can use the following function to segment the Tumor region. The result is an `sf` polygon. Internally, the functions estimate the bandwidth of the smoothing kernel (`bndw`) and the intensity threshold (`thres`) based on the distribution showed above.

```{r day3-1-spatial-stats-19}
tumor <- reconstructShapeDensitySPE(
  sfe,
  marks = "SingleR_label",
  markSelect = "Tumor",
  imageCol = "imageName")
```

We can directly plot the resulting polygon using `geom_sf`. Note that the argument `inherit.aes = FALSE` is important for the function to work correctly.

```{r day3-1-spatial-stats-20}
plotSpatialFeature(sfe, features = "SingleR_label", colGeometryName = "centroids") +
  guides(col = guide_legend(override.aes = list(size = 3))) +
  geom_sf(data = tumor, fill = NA, color = "black", linewidth = 0.75,
          inherit.aes = FALSE) +
  scale_color_manual(values = ctc)
```
::: callout-important

### Exercise
Do you think this is a good reconstruction / segmentation?
:::

::: {.callout-tip collapse="true"}
### Answer
Overall the segmentation seems to have worked fine. However, we note a few things:
1. There is one very small regions with a few Tumor cells present.
2. There might be some under-segmentation, i.e., there are some Myeloid and Endothelial cells included in the Tumor regions. As they are part of the Tumor micro-environment we could also include them in our segmentation / defintion of tumor regions.
In general, the *correct* segmentation depends heavily on the research question. Using `sosta` we have a data driven way of segmentation that is not influenced by subjective annotation bias.
:::

### Cell-based analysis

For later analyses, we will assign the cells to each tumor region.

```{r day3-1-spatial-stats-21}
# Define unique colnames by numbering the cells
colnames(sfe) <- paste0("cell_", c(1:dim(sfe)[2]))
assign <- assingCellsToStructures(sfe, tumor, imageCol = "imageName")

# Assign using the correct order of the columns in the spe object
sfe$tumorAssign <- assign[colnames(sfe)]
```

### Distance to tumor border

We can now find the distance to the closest structure / tumor boundary using the function `minBoundaryDistances`. As the function is defined to work with multiple samples, we have to give the `imageCol` that we defined above.

```{r day3-1-spatial-stats-22}
sfe$minDist <- minBoundaryDistances(
    spe = sfe, imageCol = "imageName",
    structColumn = "tumorAssign", allStructs = tumor
)
```

Again, we can plot the result. Note that we have boundary effects due to the FOV, cells that are close to the field-of-view boundary might actually be deep inside the tumor.

```{r day3-1-spatial-stats-23}
plotSpatialFeature(sfe, features="minDist", colGeometryName="centroids") +
  scale_colour_gradient2(low = "darkred", high = "navy",)
```
We can now visualize the density distribution with respect to the tumor borders in a violin plot. Negative values correspond to cells within a tumor structure. 

```{r day3-1-spatial-stats-24}
colData(sfe) |>
  as.data.frame() |>
  mutate(SingleR_label = reorder(SingleR_label, minDist, FUN = median)) |>
  ggplot(aes(x = minDist, y = SingleR_label, fill = SingleR_label)) +
  geom_violin() +
  theme_bw() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = ctc)
```
::: callout-important
### Exercise
What do you notice?
:::

::: {.callout-tip collapse="true"}
### Answer
As expected, almost all of the tumor cells have negative distances (within the tumor structures). There are some Myeloid, Endothelial and T cells within the tumor regions and in the near vicinity. This hints at the composition of the tumor microenvironment. As outlined above, the values depend on the exact segmentation and definition of the tumor region.
:::

### Cell type proportions

Next, we look at the absolute and relative cell type proportions in each tumor region.

```{r day3-1-spatial-stats-25}
sel <- colData(sfe) |>
  as.data.frame() |>
  count(tumorAssign) |>
  filter(!is.na(tumorAssign), n > 50) |> # select regions with at least 50 cells
  pull(tumorAssign)
```


```{r day3-1-spatial-stats-26}
df <- colData(sfe) |>
  as.data.frame() |>
  filter(tumorAssign %in% sel) |>
  count(tumorAssign, SingleR_label)

p0 <- ggplot(df, aes(reorder(tumorAssign, n),
                     x = n, fill = SingleR_label)) +
  theme_bw() +
  scale_fill_manual(values = ctc)

p1 <- p0 +
  geom_col() +
  labs(x = "# cells", y = "region")

p2 <- p0 +
  geom_col(position = "fill") +
  labs(x = "% cells", y = "region") +
  scale_x_continuous(labels = scales::percent)

patchwork::wrap_plots(p1, p2, guides = "collect")
```

Parts of the multi-cellular section were adapted from the following [vignette](https://helenalc.github.io/SpatialData.demo/code/csama.html).

## Additional Material

::: {.callout-note collapse="true"}
# Additional Material: Multi-sample analysis

This chapter so far showed how to perform lattice-data analysis for one sample. Lattice data analysis is not yet very common in multi-sample analyses.  One option is to compute a measure of global spatial association giving one numerical value per field-of-view. First, we will load 4 slides from the dataset (2 healthy slides and 2 cancerous slides). 

```{r day3-1-spatial-stats-27}
sfeMult <- readObject("data/day3/03.1_sfeMultNorm")
```

The loaded data is already preprocessed. The chunk below shows how this was done, you don't have to rerun it.

```{r day3-1-spatial-stats-28}
#| eval: false
#| code-fold: true
#| code-summary: "Don't run"
# identify mitochondrial genes
gn <- rowData(sfeMult)$Symbol
mt <- grepl("^MT-", gn, ignore.case = TRUE)
table(mt)
# remove them
sfeMult <- sfeMult[!mt, ]

#calculate the size factors
sf <- centerSizeFactors(colSums(assay(sfeMult)))
#normalise the data
normed <- scrapper::normalizeCounts(assay(sfeMult), size.factors = sf)
logcounts(sfeMult) <- normed
```

Next, we will define a neighbourhood on which we want to compute spatial statistics metrics. For computational reasons, we will consider a $k$ nearest neighbourhood around the cells.

```{r day3-1-spatial-stats-29}
allGraphs <- findSpatialNeighbors(sfeMult,
    method = "knearneigh", # wraps the spdep function with the same name
    k = 5,
    zero.policy = TRUE,
    sample_id = "all"
  )

colGraphs(sfeMult, "knn5", sample_id = "all") <- allGraphs
sfeMult
```

::: callout-important
## Question

- Given you have now a `SpatialFeatureExperiment` object with four samples, how would you compare a global indicator of spatial association across conditions, possibly with multiple samples? As an example, calculate global univariate Moran's $I$ for the gene "PIGR".

- An important flag in this question is `sample_id = "all"`. For reasons of having isolated cell patches, consider a $k$-NN neighbourhood this time instead of `findVisiumGraph`.

:::

::: {.callout-tip collapse="true"}
## Answer

First, we need to calculate univariate Moran's $I$ for all samples

```{r day3-1-spatial-stats-30}
sfeMult <- runUnivariate(sfeMult,
                     features = "PIGR",
                     colGraphName = "knn5",
                     exprs_values = "logcounts",
                     sample_id = "all",
                     type = "moran.mc",
                     nsim = 200)
```

```{r day3-1-spatial-stats-31}
res <- rowData(sfeMult)["PIGR",]
res
```

In order to model these measures, we have to convert the data into long format

```{r day3-1-spatial-stats-32}
df_long <- res |> as.data.frame() |>
  tibble::rownames_to_column("gene") |>
  pivot_longer(
    cols = starts_with("moran.mc_statistic"),
    names_to = c("measure", "sample"),
    names_pattern = "(.*)statistic_(.*)",
    values_to = "morans"
  ) |> separate(sample, into = c("condition", "replicate"), sep = "_") |> 
  select(c( "gene", "condition",  "replicate", "measure", "morans"))

df_long
```

```{r day3-1-spatial-stats-33}
table(df_long$condition, df_long$replicate)
```

We see that we have a nested variance structure in this data. We have two conditions "Normal" and "Cancer". We note that the patient identifiers are not unique, meaning that we have multiple samples from the same patient.

In order to account for this dependence, we have to model the data with a mixed-effects model. One option is the `lme4` package in R.

```{r day3-1-spatial-stats-34}
#convert condition to a factor and relevel to have Normal as comparison group
df_long$condition <- as.factor(df_long$condition) |>
  relevel(ref = "Normal")
mdl <- lmer(morans ~ condition + (1|replicate), data = df_long)

summary(mdl)
```

We note that in this example there is no difference in the global Moran's $I$ value for the gene "PIGR". This not surprising since a model with 3 independent samples is strongly under-powered.

An important step in interpreting a statistical model is looking at some model diagnostics. We will plot the QQ plot of this mixed effects model. We note that the four samples follow the theoretical line well. But we notice again, that having only four samples is a great limitation of this dataset.

```{r day3-1-spatial-stats-35}
qqnorm(resid(mdl), pch = 16)
qqline(resid(mdl))
```

:::
:::

::: {.callout-note collapse="true"}
# Additional Material: Gene expression within structures

Now we will subset to only the tumor regions.

```{r day3-1-spatial-stats-36}
sfe_sub <- sfe[, sfe$tumorAssign %in% sel]
```

We will pseudo-bulk within each tumor and do PCA.

```{r day3-1-spatial-stats-37}
# pseudo-bulk at the structure level
agg <- scrapper::aggregateAcrossCells(counts(sfe_sub), list(groups = sfe_sub$tumorAssign))
# normalize counts
nc <- normalizeCounts(agg$sums, size.factors = colSums(agg$sums))
# run PCA
pcaTum <- runPca(nc)
```

Let's look at the three top features (genes) for each principle component.
```{r day3-1-spatial-stats-38}
# Top features for PCs 1 to 3
apply(pcaTum$rotation[, 1:3], 2, \(.) {
    . <- sort(., decreasing=TRUE)
    head(names(.), 10)
})
```

Let's look at some features in the subset and the whole FOV.

```{r day3-1-spatial-stats-39}
#| fig-width: 16
plotSpatialFeature(sfe_sub, c("COL3A1", "FTH1", "PYGB"), ncol=3)
```

```{r day3-1-spatial-stats-40}
#| fig-width: 16
plotSpatialFeature(sfe, c("COL3A1", "FTH1", "PYGB"), ncol=3)
```

::: callout-important
### Exercise
What do you notice?
:::

::: {.callout-tip collapse="true"}
### Answer
Interestingly, several top-loaded genes for these PCs are strongly associated with non-tumor cell types. This suggests that the PCA is capturing variance that is driven by non-tumor cells within the tumor regions. Because the expression to these cells deviate so drastically from the dominant pseudo-bulk tumor profile, PCA identifies them as an important axis of variation. This could either result from the over-segementation discussed above or indicate some interesting biology or mixing patterns at the border (as we have seen in the distance density plots). This would be interesting to investigate further, e.g., check which cell types dominantly express these markers.
:::

::: callout-important
### Bonus Exercise
Rerun the segmentation with a higher intensity threshold and rerun the PCA analysis. Do you notice a change in the top genes?
:::

::: {.callout-tip collapse="true"}
### Answer
To generate a more refined segmentation, the intensity threshold in `reconstructShapeDensitySPE` has to be increased to around `2.75E-4`. The rest follows as above.

```{r day3-1-spatial-stats-41}
#| eval: false
reconstructShapeDensitySPE(
  sfe,
  marks = "SingleR_label",
  markSelect = "Tumor",
  imageCol = "imageName",
  thres = 2.75E-4
)
```

You should see some change, but the non-tumor genes are still picked up. This further indicates that we might see an interesting biological pattern.
:::
:::

```{r day3-1-spatial-stats-42}
#| eval: false
#| code-fold: true
#| code-summary: "Save the object"
alabaster.sfe::saveObject(sfe, "results/day3/03.1_sfe")
```

