---
title: "Exercise 4A: Visium HD Normalization"
format: html
editor: source
editor_options:
  chunk_output_type: console
execute:
  warning: false
---

## Normalization and scaling for Visium HD

In this exercise, we normalize the filtered Visium HD object from Exercise 2. The observations are 16 um spatial bins, not individual cells.

Standard library-size normalization is a useful starting point, but spatial data need careful interpretation: total counts can vary because of local capture efficiency, tissue density, morphology, or real biology.

## Learning objectives

By the end of this exercise, you will be able to:

- Add log-normalized values to a Visium HD `SpatialExperiment`.
- Inspect size factors and assay names.
- Scale selected features for visualization and downstream multivariate methods.
- Discuss why spatially-aware normalization can be useful.

## Libraries

```{r}
#| message: false
#| warning: false
#| output: false

library(SpatialExperiment)
library(scuttle)
library(scater)
library(HDF5Array)
library(ggspavis)
```

## Load the Visium HD object

We start with the filtered Visium HD object saved at the end of Exercise 2.

```{r}
spe <- loadHDF5SummarizedExperiment(dir = "results/day1", prefix = "01.2_spe")
spe
```

## Log-normalization

The object contains raw counts. We use `scuttle::logNormCounts()` to compute size factors and add a `logcounts` assay.

```{r}
assayNames(spe)

spe <- logNormCounts(spe)

assayNames(spe)
summary(sizeFactors(spe))
```

::: callout-important
## Exercise 1

What assay was added to the Visium HD object? What do the size factors represent?
:::

::: {.callout-tip collapse="true"}
## Answer

```{r}
assayNames(spe)
summary(sizeFactors(spe))
```

The `logcounts` assay contains log-normalized expression values. The size factors adjust for differences in total count depth between bins.
:::

We can inspect whether size factors have a spatial pattern.

```{r}
colData(spe)$sizeFactor <- sizeFactors(spe)

plotVisium(
  spe,
  annotate = "sizeFactor",
  zoom = TRUE,
  point_shape = 22
)
```

::: callout-important
## Exercise 2

Do size factors show a spatial pattern across the tissue region? What biological or technical effects could explain this?
:::

## Scaling selected features

Normalization adjusts for differences in total counts between observations. Scaling is a different step: it centers and rescales features so that they can be compared on a common scale. This is often useful before multivariate methods such as PCA, clustering, or heatmaps.

Here we demonstrate scaling on a small set of marker genes.

```{r}
marker_genes <- c("PIGR", "CEACAM5", "MUC17", "OLFM4")
marker_genes <- intersect(marker_genes, rownames(spe))
marker_genes
```

```{r}
scale_rows <- function(x) {
  t(base::scale(t(x)))
}

visium_scaled <- scale_rows(as.matrix(logcounts(spe)[marker_genes, ]))
visium_scaled[, 1:5]
```

::: callout-important
## Exercise 3

What does a positive scaled value mean for a gene in one bin? Why should scaling be applied after normalization rather than directly to raw counts?
:::

::: {.callout-tip collapse="true"}
## Answer

A positive scaled value means that the bin has expression above the average for that gene, relative to the variation of that gene across the dataset. Scaling should be applied after normalization because raw counts are strongly affected by sequencing depth.
:::

## Optional: spatially-aware normalization

::: callout-note
## Optional discussion

Standard library-size normalization treats differences in total counts as mostly technical. In spatial transcriptomics, this assumption can be imperfect: total counts can vary because of tissue density, local capture efficiency, morphology, or real biology.

Spatially-aware normalization methods try to account for spatial structure while avoiding the removal of true spatial biological signal. This is an active area of method development. OSTA discusses these issues in the context of spatial workflows, and packages such as `SpaNorm` can be explored as optional extensions.

For this course, we keep the required workflow simple and use `logNormCounts()`. When interpreting the downstream results, remember that spatial structure in size factors or total counts may itself be informative.
:::

::: callout-important
## Exercise 4

Look back at the spatial plot of Visium HD size factors. Would you expect a spatially-aware normalization method to remove all of that pattern? Why or why not?
:::

::: {.callout-tip collapse="true"}
## Answer

Not necessarily. Some spatial pattern in size factors may reflect technical effects, but some may reflect real tissue structure, local cell density, or biological differences. A spatially-aware method should avoid removing meaningful spatial biology while reducing unwanted technical variation.
:::

## Save normalized object

We save the normalized Visium HD object for downstream exercises.

```{r}
dir.create("results/day2", showWarnings = FALSE, recursive = TRUE)

saveHDF5SummarizedExperiment(
  spe,
  dir = "results/day2",
  prefix = "02.1_spe",
  replace = TRUE,
  chunkdim = NULL,
  level = NULL,
  as.sparse = NA,
  verbose = NA
)
```

Clear your environment:

```{r}
#| eval: false
rm(list = ls())
gc()
.rs.restartR()
```

::: callout-important
**Key Takeaways:**

- `logNormCounts()` adds a `logcounts` assay and stores size factors.
- Scaling centers and rescales genes after normalization.
- Spatial structure in size factors may be technical, biological, or both.
:::
