---
title: "Exercise 6: Clustering and Domains"
format: html
editor: source
editor_options: 
  chunk_output_type: console
execute:
  warning: false
---

```{r day2-3-clustering-find-markers-1}
#| message: false
#| warning: false
#| output: false

library(Banksy)
library(igraph)
library(ggspavis)
library(BayesSpace)
library(nnSVG)
library(HDF5Array) # For loading object
library(patchwork)
library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(Voyager)
library(scran) 
library(scater)
library(bluster)
library(SingleR)
library(qs2)
library(viridis)
library(RColorBrewer)
library(pheatmap)
library(alabaster.sfe)

```


## Clustering and Spatially-aware clustering

In the next section we will explore clustering methods for spatial transcriptomics data, including classical non-spatially aware and spatially-aware approaches. 

## Learning Objectives
By the end of this exercise, you will be able to:

-  Perform non-spatially aware clustering using graph-based methods (Leiden clustering).
-  Perform spatially-aware clustering using Bayesian modeling (BayesSpace) and graph-based methods (Banksy + Leiden).
-  Visualize and compare clustering results from different methods on tissue slides.

## Clustering

We will start by loading the normalized Visium HD dataset stored in the `SpatialFeatureExperiment` object generated in the previous exercise. It contains information on the HVGs and dimensionationality reduction computed with Banksy and PCA.

```{r day2-3-clustering-find-markers-2}
# Reload the SpatialFeatureExperiment object if not in the R session already
sfe <- readObject("results/day2/02.2_sfe/")
sfe
```

### Non-spatial aware clustering

Clustering methods can be categorized into non-spatially aware, e.g., these include the  methods classically used for scRNA-seq analysis, and spatially-aware methods. We will first explore non-spatially aware clustering using graph-based methods. 

As commonly done in single-cell RNA-seq analysis, we can perform graph-based clustering using the Leiden algorithm on a shared nearest-neighbor (SNN) graph constructed from the PCA results.

```{r day2-3-clustering-find-markers-3}
# build cellular shared nearest-neighbor (SNN) graph
g <- makeSNNGraph(reducedDim(sfe, "PCA"), type="jaccard", k=20)
# cluster using Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=0.6)
# assign cluster labels to sfe object
sfe$Leiden <- factor(k$membership)
table(sfe$Leiden)
```


### Spatialy aware clustering

Clustering methods can also incorporate spatial information to identify spatial domains or regions in the tissue. 

#### Probabilistic: BayesSpace

The `BayesSpace` package performs spatially-aware clustering using a Bayesian modeling approach. Unfortunately it is not possible yet to use it on segmented data, see for example: https://github.com/edward130603/BayesSpace/issues/136 

#### Exercise 1: Bonus 

Go back to our binned data stored in a `SpatialExperiment` object to run `BayesSpace`. We need to quickly go through all the steps of normalization, HVG selection and dimensionality reduction.

```{r day2-3-clustering-find-markers-4}
spe <- loadHDF5SummarizedExperiment(dir="results/day1/", prefix="01.1a_spe_")
spe <- logNormCounts(spe)

dec <- modelGeneVarByPoisson(spe)
dec <- dec[order(dec$bio, decreasing = T), ]
hvg <- getTopHVGs(dec, n = 1000) 
rowData(spe)$hvg <- rowData(spe)$Symbol %in% hvg

## non-spatial PCA
spe <- runPCA(spe, ncomponents = 10, subset_row = rowData(spe)$hvg)

# Banksy PCA (for comparison)
k <- 20   # consider first order neighbors
l <- 0.8 # use spatial information with weight lambda
a <- "logcounts" # assay to use
xy <- c("array_row", "array_col")  # spatial coordinate names
set.seed(123)
tmp <- computeBanksy(spe[rowData(spe)$hvg, ], 
                     assay_name=a, 
                     coord_names=xy, 
                     k_geom=k, 
                     parallel=T,
                     num_cores=4)
tmp <- runBanksyPCA(tmp, 
                    lambda=l, 
                    npcs=10)
reducedDim(spe, "PCA_banksy") <- reducedDim(tmp, "PCA_M0_lam0.8") 


# Prepare data for 'BayesSpace'
# skipping PCA (already computed)
spe <- spatialPreprocess(spe, skip.PCA=TRUE)

# perform spatial clustering with 'BayesSpace'
# using 'd=10' PCs and targeting 'q=8' clusters
spe <- spatialCluster(spe, q=8, d=10, nrep=1e3, burn.in=100) 
spe$BayesSpace <- factor(spe$spatial.cluster)
table(spe$BayesSpace)
```

#### Graph-based: Banksy

We have computed spatially-aware principal components using Banksy already in the previous practical. We will now used these to perform SNN graph-based Leiden clustering, in order to obtain spatially-aware clusters.
When building the SNN graph, we will use the same parameters we have used before for the non-spatially aware graph.

```{r day2-3-clustering-find-markers-5}
# perform SNN graph-based clustering on 'Banksy' PCs using
g <- buildSNNGraph(sfe, use.dimred="PCA_banksy", type="jaccard", k=20)
# cluster using Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=0.2)
sfe$Banksy <- factor(k$membership)
table(sfe$Banksy)
```

#### Exercise 2: Bonus 
Perform a similar graph-based clustering on Banksy PCs in the binned `spe` object

## Comparison of results from the different clustering methods

Let's visualize the clustering results from the different methods on the tissue slide.

```{r day2-3-clustering-find-markers-6}
## 2 methods to compare for the sfe object
ks <- c("Leiden", "Banksy") 
plotSpatialFeature(sfe, features = ks)

## If BayesSpace was run on the spe object
# Colorblind friendly palette
data("ditto_colors")
plotVisium(spe,
           annotate = "BayesSpace",
           zoom = TRUE,
           image = FALSE,
           point_size = 1.5,
           point_shape = 22) + 
 scale_fill_manual(values = ditto_colors)
```

It can help us to visualise the clusters one by one, in order to better see their spatial distribution. We take as an example the Banksy clusters:

```{r day2-3-clustering-find-markers-7}
# plot selected clusters in order of frequency, highlighting cells assigned to cluster 'k'
lapply(names(sort(table(sfe$Banksy), decreasing = TRUE)), \(k) {
    sfe$foo <- sfe$Banksy == k
    sfe <- sfe[, order(sfe$foo)]
    plt <- plotSpatialFeature(sfe, features = "foo")
    plt + ggtitle(k) + guides(fill="none")
}) |>
  wrap_plots(nrow=2) &
  scale_color_manual(values=c("white", "red")) &
  theme(plot.title=element_text(hjust=0.5), legend.position="none")
```


:::{.callout-important}
### Exercise 3
Visualise the clusters obtained with the 3 methods, and compare their spatial distribution. What do you observe?

What is the interpretation of a domain compared to a cluster?

:::   

<!-- 
TO DO: answer this
-->

## Save the object

Save the `SpatialFeatureExperiment` object for the next steps:

```{r day2-3-clustering-find-markers-8}
alabaster.sfe::saveObject(sfe, "results/day2/02.3_sfe")
```

Potentially save the `SpatialExperiment` object used in bonus exercises

```{r day2-3-clustering-find-markers-9}
saveHDF5SummarizedExperiment(spe,
  dir = "results/day2", prefix = "02.3_spe_", replace = TRUE,
  chunkdim = NULL, level = NULL, as.sparse = NA,
  verbose = NA
)
```

Clear your environment:
```{r day2-3-clustering-find-markers-10}
#| eval: false
rm(list = ls())
gc()
.rs.restartR()
```
